A Bayesian approach to uncertainty in word embedding bias estimation

### 2023-06-15 paper/paperDraft.tex
paper sent to arxiv

It has been suggested\footnote{See for instance {[}1,4,5,8,11,12{]}.}
that language models can learn implicit biases that reflect harmful
stereotypical thinking---for example, the (vector corresponding to the)
word \textit{she} might be much closer in the vector space to the word
\textit{cooking} than the word \textit{he}. Such phenomena are
undesirable at least in some downstream tasks, such as web search,
recommendations, and so on. To investigate such issues, several measures
of bias in word embeddings have been formulated and applied. Our goal is
to use two prominent examples of such measures to argue that this
approach oversimplifies the situation and to develop a Bayesian
alternative.

A common approach in natural language processing is to represent words
by vectors of real numbers---such representations are called
\emph{embeddings}. One way to construct an embedding---we will focus our
attention on non-contextual language models\footnote{One example of a
  contextualized representation is BERT. Another is GPT.}---is to use a
large corpus to train a neural network to assign vectors to words in a
way that optimizes for co-occurrence prediction accuracy. Such vectors
can then be compared in terms of their similarity---the usual measure is
cosine similarity---and the results of such comparisons can be used in
downstream tasks. Roughly speaking, cosine similarity is an imperfect
mathematical proxy for semantic similarity {[}15{]}.

One response to the raising of the issue of bias in natural language
models might be to say that there is not much point to reflecting on
such biases, as they are unavoidable. This unavoidability might seem in
line with the arguments to the effect that learning algorithms are
always value-laden {[}9{]}: they employ inductive methods that require
design-, data-, or risk-related decisions that have to be guided by
extra-algorithmic considerations. Such choices necessarily involve value
judgments and have to do, for instance, with what simplifications or
risks one finds acceptable. Admittedly, algorithmic decision making
cannot fulfill the value-free ideal, but this only means that even more
attention needs to be paid to the values underlying different techniques
and decisions, and to the values being pursued in a particular use of an
algorithm.

Another response might be to insist that there is no bias introduced by
the use of machine learning methods here, since the algorithm is simply
learning to correctly predict co-occurrences based on what ``reality''
looks like. However, this objection overlooks the fact that we, humans,
are the ones who construct this linguistic reality, which is shaped in
part by the natural language processing tools we use on a massive scale.
Sure, if there is unfairness and our goal is to diagnose it, we should
do complete justice to learning it in the model used to study it. One
example of this approach is {[}3{]}, where the authors use language
models to study the shape of certain biases across a century.

However, if our goal is to develop downstream tools that perform tasks
that we care about without further perpetuating or exacerbating harmful
stereotypes, we still have good reasons to try to minimize the negative
impact. Moreover, it is often not the case that the corpora mirror
reality---to give a trivial example, heads are spoken of more often than
kidneys, but this does not mean that kidneys occur much less often in
reality than heads. To give a more relevant example, the
disproportionate association of female words with female occupations in
a corpus actually greatly exaggerates the actual lower disproportion in
the real distribution of occupations {[}6{]}.

In what follows, we focus on two popular measures of bias applicable to
many existing word embeddings, such as
\emph{GoogleNews},\footnote{GoogleNews-vectors-negative300, available at  \url{https://github.com/mmihaltz/word2vec-GoogleNews-vectors}.}
\emph{Glove}\footnote{Available at \url{https://nlp.stanford.edu/projects/glove/}.}
and
\emph{Reddit Corpus}\footnote{Reddit-L2 corpus, available at  \url{http://cl.haifa.ac.il/projects/L2/}.}:
\emph{Word Embedding Association Test} (\textsf{WEAT}) {[}8{]}, and
\emph{Mean Average Cosine Distance} (\textsf{MAC}) {[}12{]}. We first
explain how these measures are supposed to work. Then we argue that they
are problematic for various reasons---the key one being that by
pre-averaging data they manufacture false confidence, which we
illustrate in terms of simulations showing that the measures often
suggest the existence of bias even if by design it is non-existent in a
simulated dataset.

@@

Most of the problems that we point out generalize to any existing
approach that focuses on chasing a single numeric metric of bias. (1)
They treat the results of pre-averaging as raw data in statistical
significance tests, which in this context is bound to overestimate
significance. We show similar results can easily be obtained when
sampling from null models with no bias. (2) The word list sizes and
sample sizes used in the studies are usually small,\footnote{Depending
  on a list for {[}8{]} the range for protected words is between 13 and
  100, and for attributes between 16 and 25; for {[}12{]} the range for
  protected words is between 14 and 18, and for attributes between 11
  and 25.} (3) Many studies do not use any control predicates, such as
random neutral words or neutral human predicates for comparison.

@@

The underlying intuition is that if a particular harmful stereotype is
learned in a particular embedding, then certain groups of words will be
systematically closer to (or further from) each other. This gives rise
to the idea of protected groups---for example, in guiding online search
completion or recommendation, female words might require protection in
that they should not be systematically closer to stereotypically female
job names, such as ``nurse'', ``librarian'', ``waitress'', and male
words require protection in that they should not be systematically
closer to toxic masculinity stereotypes, such as ``tough'', ``never
complaining'' or ``macho''.\footnote{However, for some research-related
  purposes, such as the study of stereotypes across history {[}3{]},
  embeddings which do not protect certain classes may also be useful.}

@@

The use of projections in bias estimation has been criticized for
instance in {[}5{]}, where it is pointed out that while a higher average
similarity to the gender direction might be an indicator of bias with
respect to a given class of words, it is only one possible manifestation
of it, and reducing the cosine similarity to such a projection may not
be sufficient to eliminate bias. For instance, ``math'' and ``delicate''
might be equally similar to a pair of opposed explicitly gendered words
(\emph{she}, \emph{he}), while being closer to quite different
stereotypical attribute words (such as \emph{scientific} or
\emph{caring}). Further, it is observed in {[}5{]} that most word pairs
retain similarity under debiasing meant to minimize projection-based
bias.\footnote{In {[}1{]} another method which involves analogies and
  their evaluations by human users on Mechanical Turk is also used. We
  do not discuss this method in this paper, see its criticism in
  {[}17{]}.}

A measure of bias in word embeddings which does not proceed by
identifying bias directions (such as a gender vector), the Word
Embedding Association Test (\textsf{WEAT}), has been proposed in
{[}8{]}. The idea here is that the bias between two sets of target
words, \(X\) and \(Y\) (we call them protected words), should be
quantified in terms of the cosine similarity between the protected words
and attribute words coming from two sets of stereotype attribute words,
\(A\) and \(B\) (we will call them attributes). For instance, \(X\)
might be a set of male names, \(Y\) a set of female names, \(A\) might
contain stereotypically male-related, and \(B\) stereotypically
female-related career words. The association difference for a particular
word \(t\) (belonging to either \(X\) or \(Y\)) is:

@@

\textsf{WEAT} is inspired by the Implicit Association Test (IAT)
{[}18{]} used in psychology, and in some applications it uses almost the
same word sets, allowing for a \emph{prima facie} sensible comparison
with bias in humans. In {[}8{]} the authors argue that significant
biases---thus measured--- similar to the ones discovered by IAT can be
discovered in word embeddings. In {[}11{]} the methodology is extended
to a multilingual and cross-lingual setting, arguing that using
Euclidean distance instead of cosine similarity does not make much
difference, while the bias effects vary greatly across embedding
models.\footnote{Interestingly, with social media-text trained
  embeddings being less biased than those based on Wikipedia.} A similar
methodology is employed in {[}4{]}. The authors employ word embeddings
trained on corpora from different decades to study the shifts in various
biases through the century.\footnote{Strictly speaking, these authors
  use Euclidean distances and their differences, but the way they take
  averages and averages thereof is analogous, and so what we will have
  to say about pre-averaging leading to false confidence applies to this
  methodology as well.}

@@

\noindent The \textsf{WEAT} scores reported by {[}8{]} for lists of
words for which the embeddings are supposedly biased range from 2.06 to
1.81, and the reported \(p\)-values are in the range of
\(10^{-7}-10^{-2}\) with one exception for \emph{Math vs Arts}, where it
is \(.018\).

@@

Imagine there are two groups of protected words, each of size 8, and two
groups of stereotypical attributes, of the same size.\footnote{16 is the
  sample size used in the WEAT7 word list, which is not much different
  from the other sample sizes in word lists used by {[}8{]}).} Each such
a collection of samples, as far as our question is involved, is
equivalent to a sample of \(16^2\) cosine distances. Further, imagine
there really is no difference between these groups of word and the model
is in fact null. That is, we draw the cosine distances from the
\(\mathsf{Normal}(0,.08)\) distribution.\footnote{\(.08\) is
  approximately the empirical standard deviation observed in fairly
  large samples of neutral words.}

@@

The difficulties are exacerbated by the fact that statistical tests are
based on bootstrapping from a relatively small data sets, which is quite
likely to underestimate the population variance. To make our point
clear, let us avoid bootstrapping and work with the null generative
model with \(\mathsf{Norm}(0,.08)\) for both word groups. We keep the
sizes the same: we have eight protected words in each group, sixteen in
total, and for each we randomly draw 8 distances from hypothetical \(A\)
attributes, and \(8\) distances from hypothetical \(B\) attributes.
Calculate the test statistic and effect size the way {[}8{]} did. Do
this 10000 times, each time calculating \textsf{WEAT} and \textsf{s}
values, and look at what the distributions of these values are on the
assumption of the null model with realistic empirically motivated raw
data point standard deviation.

@@

Bayesian data analysis takes prior probability distributions, a
mathematical model structure and the data, and returns the posterior
probability distributions over the parameters of interest, thus
capturing our uncertainty about their actual values. One important
difference between such a result and the result of a classical
statistical analysis is that classical confidence intervals (CIs) have a
rather complicated and somewhat confusing interpretation, which has
little to do with the posterior probability distribution.\footnote{Here
  are a few usual problems. CIs are often mistakenly interpreted as
  providing the probability that a resulting confidence interval
  contains the true value of a parameter. CIs bring confusion also with
  regard to precision, it is a common mistake to interpret narrow
  intervals as the ones corresponding to a more precise knowledge.
  Another fallacy is to associate CIs with likelihood and stating that
  values within a given interval are more probable than the ones outside
  it. The theory of confidence intervals does not support the above
  interpretations. CIs should be plainly interpreted as a result of
  certain procedure (there are many ways to obtain CIs from a given set
  of data) that will in the long run contain the true value if the
  procedure is performed a fixed amount of times. For a nice survey and
  explanation of these misinterpretations, see {[}16{]}. For a
  psychological study of the occurrence of such misinterpretations, see
  REF. In this study, 120 researchers and 442 students were asked to
  assess the truth value of six false statements involving different
  interpretations of a CI. Both researchers and students endorsed, on
  average, more than three of these statements. See also {[}7{]} for a
  list of problems with the classical approach.}

@@

{[}20{]} employ Bayesian methods to estimate uncertainty in NLP tasks,
but they apply their Bayesian Neural Networks-based method to sentiment
analysis or named entity recognition, not to bias. {[}2{]} correctly
argues that bias estimate should not be expressed as a single number
without taking into account that the estimate is made using a sample of
data and therefore has intrinsic uncertainty. The authors suggest to use
Bernstein bounds to gauge the uncertainty in terms of confidence
intervals. We do not discuss this approach extensively, as we think that
confidence intervals are quite problematic for several reasons, among
others the confusing interpretability. We do not think that Bernstein
bounds provide the best solution to the problem. Applying this method to
a popular WinoBias dataset leads to the conclusion that more than 11903
samples are needed to claim a 95\% confidence interval for a bias
estimate. This amount vastly exceeds the existing word lists for bias
estimation. We propose a more realistic Bayesian method. Our conclusion
is still that the word lists are sometimes too small, but at least
allows for gauging uncertainty as we go on to improve our methodology
and extend the lists gradually.

{[}19{]} criticize some exisisting bias metrics such as \textsf{MAC} or
\textsf{WEAT} on the grounds of them not satisfying some general formal
principles, such as magnitude-comparability, and they propose a
modification, called \textsf{SAME}. {[}13{]} develop a generalization of
\textsf{WEAT} meant to apply to sets of sentences, \textsf{SEAT}, which
basically applies \textsf{WEAT} to vector representations of sentences.
The authors, however, still pre-average and play the game of finding a
single-number metric, so our remarks apply.

\todo{More references to similar papers}

A Bayesian data analysis with hierarchical models of cosine distances
between protected words, control group words, and stereotypical
attributes provides more modest and realistic assessment of the
uncertainty involved. It reveals that much complexity is hidden when one
instead chases single bias metrics present in the literature.

After introducing the method, we apply it to multiple word embeddings
and results of supposed debiasing, putting forward some general
observations that are not exactly in line with the usual picture painted
in terms of \textsf{WEAT} or \textsf{MAC} (and the problem generalizes
to any approach that focuses on chasing a single numeric metric): the
word list sizes and sample sizes used in the studies are usually small.
posterior density intervals are fairly wide, often the differences
between associated, different and neutral human predicates, are not very
impressive. Also, a preliminary inspection suggests that the
desirability of changes obtained by the usual debiasing methods is
debatable. The tools that we propose, however, allow for a more
fine-grained and multi-level evaluation of bias and debiasing in
language models without losing modesty about the uncertainties involved.

The short general lesson here is this: it's complicated. Instead of
pretending it isn't and chasing single-number metrics, we should rather
devote attention to more nuanced and methodologically sound analysis.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Ethayarajh2020Bernstein}{}}%
\CSLLeftMargin{{[}2{]} }%
\CSLRightInline{Kawin Ethayarajh. 2020. Is your classifier actually
biased? Measuring fairness under uncertainty with bernstein bounds.
\emph{CoRR} abs/2004.12332, (2020). Retrieved from
\url{https://arxiv.org/abs/2004.12332}}

\leavevmode\vadjust pre{\hypertarget{ref-Garg2017hundredYears}{}}%
\CSLLeftMargin{{[}3{]} }%
\CSLRightInline{Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James
Zou. 2017. Word embeddings quantify 100 years of gender and ethnic
stereotypes. \emph{Proceedings of the National Academy of Sciences} 115,
(November 2017).
DOI:https://doi.org/\href{https://doi.org/10.1073/pnas.1720347115}{10.1073/pnas.1720347115}}

\leavevmode\vadjust pre{\hypertarget{ref-Garg2018years}{}}%
\CSLLeftMargin{{[}4{]} }%
\CSLRightInline{Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James
Zou. 2018. Word embeddings quantify 100 years of gender and ethnic
stereotypes. \emph{Proceedings of the National Academy of Sciences} 115,
16 (April 2018), E3635--E3644.
DOI:https://doi.org/\href{https://doi.org/10.1073/pnas.1720347115}{10.1073/pnas.1720347115}}

\leavevmode\vadjust pre{\hypertarget{ref-Gonen2019lipstick}{}}%
\CSLLeftMargin{{[}5{]} }%
\CSLRightInline{Hila Gonen and Yoav Goldberg. 2019. Lipstick on a pig:
{D}ebiasing methods cover up systematic gender biases in word embeddings
but do not remove them. In \emph{Proceedings of the 2019 conference of
the north {A}merican chapter of the association for computational
linguistics: Human language technologies, volume 1 (long and short
papers)}, Association for Computational Linguistics, Minneapolis,
Minnesota, 609--614.
DOI:https://doi.org/\href{https://doi.org/10.18653/v1/N19-1061}{10.18653/v1/N19-1061}}

\leavevmode\vadjust pre{\hypertarget{ref-gordon2012reporting}{}}%
\CSLLeftMargin{{[}6{]} }%
\CSLRightInline{Jonathan Gordon and Benjamin Durme. 2013. Reporting bias
and knowledge acquisition. In \emph{AKBC 2013 - Proceedings of the 2013
Workshop on Automated Knowledge Base Construction, Co-located with CIKM
2013}, 25--30.
DOI:https://doi.org/\href{https://doi.org/10.1145/2509558.2509563}{10.1145/2509558.2509563}}

\leavevmode\vadjust pre{\hypertarget{ref-Hoekstra2014Misinterpretation}{}}%
\CSLLeftMargin{{[}7{]} }%
\CSLRightInline{Rink Hoekstra, Richard D. Morey, Jeffrey N. Rouder, and
Eric-Jan Wagenmakers. 2014. Robust misinterpretation of confidence
intervals. \emph{Psychonomic Bulletin \& Review} 21, 5 (October 2014),
1157--1164.
DOI:https://doi.org/\href{https://doi.org/10.3758/s13423-013-0572-3}{10.3758/s13423-013-0572-3}}

\leavevmode\vadjust pre{\hypertarget{ref-Caliskan2017semanticsBiases}{}}%
\CSLLeftMargin{{[}8{]} }%
\CSLRightInline{Aylin Caliskan Islam, Joanna J. Bryson, and Arvind
Narayanan. 2016. Semantics derived automatically from language corpora
necessarily contain human biases. \emph{CoRR} abs/1608.07187, (2016).
Retrieved from \url{http://arxiv.org/abs/1608.07187}}

\leavevmode\vadjust pre{\hypertarget{ref-JohnsonValueFree}{}}%
\CSLLeftMargin{{[}9{]} }%
\CSLRightInline{Gabbrielle Johnson. forthcoming. Are algorithms
value-free? Feminist theoretical virtues in machine learning.
\emph{Journal Moral Philosophy} (forthcoming).}

\leavevmode\vadjust pre{\hypertarget{ref-kruschke2015bayesian}{}}%
\CSLLeftMargin{{[}10{]} }%
\CSLRightInline{John Kruschke. 2015. \emph{Doing bayesian data analysis
(second edition)}. Academic Press, Boston.}

>>>>

It has been suggested\footnote{See for instance {[}1,2,5,6,12,13{]}.}
that language models can learn implicit biases that reflect harmful
stereotypical thinking---for example, the (vector corresponding to the)
word \textit{she} might be much closer in the vector space to the word
\textit{cooking} than the word \textit{he}. Such phenomena are
undesirable at least in some downstream tasks, such as web search,
recommendations, and so on. To investigate such issues, several measures
of bias in word embeddings have been formulated and applied. Our goal is
to use two prominent examples of such measures to argue that this
approach oversimplifies the situation and to develop a Bayesian
alternative.

A common approach in natural language processing is to represent words
by vectors of real numbers---such representations are called
\emph{embeddings}. One way to construct an embedding---we will focus our
attention on non-contextual language models\footnote{One example of a
  contextualized representation is BERT. Another is GPT.}---is to use a
large corpus to train a neural network to assign vectors to words in a
way that optimizes for co-occurrence prediction accuracy. Such vectors
can then be compared in terms of their similarity---the usual measure is
cosine similarity---and the results of such comparisons can be used in
downstream tasks. Roughly speaking, cosine similarity is an imperfect
mathematical proxy for semantic similarity {[}16{]}.

One response to the raising of the issue of bias in natural language
models might be to say that there is not much point to reflecting on
such biases, as they are unavoidable. This unavoidability might seem in
line with the arguments to the effect that learning algorithms are
always value-laden {[}10{]}: they employ inductive methods that require
design-, data-, or risk-related decisions that have to be guided by
extra-algorithmic considerations. Such choices necessarily involve value
judgments and have to do, for instance, with what simplifications or
risks one finds acceptable. Admittedly, algorithmic decision making
cannot fulfill the value-free ideal, but this only means that even more
attention needs to be paid to the values underlying different techniques
and decisions, and to the values being pursued in a particular use of an
algorithm.

Another response might be to insist that there is no bias introduced by
the use of machine learning methods here, since the algorithm is simply
learning to correctly predict co-occurrences based on what ``reality''
looks like. However, this objection overlooks the fact that we, humans,
are the ones who construct this linguistic reality, which is shaped in
part by the natural language processing tools we use on a massive scale.
Sure, if there is unfairness and our goal is to diagnose it, we should
do complete justice to learning it in the model used to study it. One
example of this approach is {[}4{]}, where the authors use language
models to study the shape of certain biases across a century.

However, if our goal is to develop downstream tools that perform tasks
that we care about without further perpetuating or exacerbating harmful
stereotypes, we still have good reasons to try to minimize the negative
impact. Moreover, it is often not the case that the corpora mirror
reality---to give a trivial example, heads are spoken of more often than
kidneys, but this does not mean that kidneys occur much less often in
reality than heads. To give a more relevant example, the
disproportionate association of female words with female occupations in
a corpus actually greatly exaggerates the actual lower disproportion in
the real distribution of occupations {[}7{]}.

In what follows, we focus on two popular measures of bias applicable to
many existing word embeddings, such as
\emph{GoogleNews},\footnote{GoogleNews-vectors-negative300, available at  \url{https://github.com/mmihaltz/word2vec-GoogleNews-vectors}.}
\emph{Glove}\footnote{Available at \url{https://nlp.stanford.edu/projects/glove/}.}
and
\emph{Reddit Corpus}\footnote{Reddit-L2 corpus, available at  \url{http://cl.haifa.ac.il/projects/L2/}.}:
\emph{Word Embedding Association Test} (\textsf{WEAT}) {[}2{]}, and
\emph{Mean Average Cosine Distance} (\textsf{MAC}) {[}13{]}. We first
explain how these measures are supposed to work. Then we argue that they
are problematic for various reasons---the key one being that by
pre-averaging data they manufacture false confidence, which we
illustrate in terms of simulations showing that the measures often
suggest the existence of bias even if by design it is non-existent in a
simulated dataset.

@@

Most of the problems that we point out generalize to any existing
approach that focuses on chasing a single numeric metric of bias. (1)
They treat the results of pre-averaging as raw data in statistical
significance tests, which in this context is bound to overestimate
significance. We show similar results can easily be obtained when
sampling from null models with no bias. (2) The word list sizes and
sample sizes used in the studies are usually small,\footnote{Depending
  on a list for {[}2{]} the range for protected words is between 13 and
  100, and for attributes between 16 and 25; for {[}13{]} the range for
  protected words is between 14 and 18, and for attributes between 11
  and 25.} (3) Many studies do not use any control predicates, such as
random neutral words or neutral human predicates for comparison.

@@

The underlying intuition is that if a particular harmful stereotype is
learned in a particular embedding, then certain groups of words will be
systematically closer to (or further from) each other. This gives rise
to the idea of protected groups---for example, in guiding online search
completion or recommendation, female words might require protection in
that they should not be systematically closer to stereotypically female
job names, such as ``nurse'', ``librarian'', ``waitress'', and male
words require protection in that they should not be systematically
closer to toxic masculinity stereotypes, such as ``tough'', ``never
complaining'' or ``macho''.\footnote{However, for some research-related
  purposes, such as the study of stereotypes across history {[}4{]},
  embeddings which do not protect certain classes may also be useful.}

@@

The use of projections in bias estimation has been criticized for
instance in {[}6{]}, where it is pointed out that while a higher average
similarity to the gender direction might be an indicator of bias with
respect to a given class of words, it is only one possible manifestation
of it, and reducing the cosine similarity to such a projection may not
be sufficient to eliminate bias. For instance, ``math'' and ``delicate''
might be equally similar to a pair of opposed explicitly gendered words
(\emph{she}, \emph{he}), while being closer to quite different
stereotypical attribute words (such as \emph{scientific} or
\emph{caring}). Further, it is observed in {[}6{]} that most word pairs
retain similarity under debiasing meant to minimize projection-based
bias.\footnote{In {[}1{]} another method which involves analogies and
  their evaluations by human users on Mechanical Turk is also used. We
  do not discuss this method in this paper, see its criticism in
  {[}18{]}.}

A measure of bias in word embeddings which does not proceed by
identifying bias directions (such as a gender vector), the Word
Embedding Association Test (\textsf{WEAT}), has been proposed in
{[}2{]}. The idea here is that the bias between two sets of target
words, \(X\) and \(Y\) (we call them protected words), should be
quantified in terms of the cosine similarity between the protected words
and attribute words coming from two sets of stereotype attribute words,
\(A\) and \(B\) (we will call them attributes). For instance, \(X\)
might be a set of male names, \(Y\) a set of female names, \(A\) might
contain stereotypically male-related, and \(B\) stereotypically
female-related career words. The association difference for a particular
word \(t\) (belonging to either \(X\) or \(Y\)) is:

@@

\textsf{WEAT} is inspired by the Implicit Association Test (IAT)
{[}19{]} used in psychology, and in some applications it uses almost the
same word sets, allowing for a \emph{prima facie} sensible comparison
with bias in humans. In {[}2{]} the authors argue that significant
biases---thus measured--- similar to the ones discovered by IAT can be
discovered in word embeddings. In {[}12{]} the methodology is extended
to a multilingual and cross-lingual setting, arguing that using
Euclidean distance instead of cosine similarity does not make much
difference, while the bias effects vary greatly across embedding
models.\footnote{Interestingly, with social media-text trained
  embeddings being less biased than those based on Wikipedia.} A similar
methodology is employed in {[}5{]}. The authors employ word embeddings
trained on corpora from different decades to study the shifts in various
biases through the century.\footnote{Strictly speaking, these authors
  use Euclidean distances and their differences, but the way they take
  averages and averages thereof is analogous, and so what we will have
  to say about pre-averaging leading to false confidence applies to this
  methodology as well.}

@@

\noindent The \textsf{WEAT} scores reported by {[}2{]} for lists of
words for which the embeddings are supposedly biased range from 2.06 to
1.81, and the reported \(p\)-values are in the range of
\(10^{-7}-10^{-2}\) with one exception for \emph{Math vs Arts}, where it
is \(.018\).

@@

Imagine there are two groups of protected words, each of size 8, and two
groups of stereotypical attributes, of the same size.\footnote{16 is the
  sample size used in the WEAT7 word list, which is not much different
  from the other sample sizes in word lists used by {[}2{]}).} Each such
a collection of samples, as far as our question is involved, is
equivalent to a sample of \(16^2\) cosine distances. Further, imagine
there really is no difference between these groups of word and the model
is in fact null. That is, we draw the cosine distances from the
\(\mathsf{Normal}(0,.08)\) distribution.\footnote{\(.08\) is
  approximately the empirical standard deviation observed in fairly
  large samples of neutral words.}

@@

The difficulties are exacerbated by the fact that statistical tests are
based on bootstrapping from a relatively small data sets, which is quite
likely to underestimate the population variance. To make our point
clear, let us avoid bootstrapping and work with the null generative
model with \(\mathsf{Norm}(0,.08)\) for both word groups. We keep the
sizes the same: we have eight protected words in each group, sixteen in
total, and for each we randomly draw 8 distances from hypothetical \(A\)
attributes, and \(8\) distances from hypothetical \(B\) attributes.
Calculate the test statistic and effect size the way {[}2{]} did. Do
this 10000 times, each time calculating \textsf{WEAT} and \textsf{s}
values, and look at what the distributions of these values are on the
assumption of the null model with realistic empirically motivated raw
data point standard deviation.

@@

Bayesian data analysis takes prior probability distributions, a
mathematical model structure and the data, and returns the posterior
probability distributions over the parameters of interest, thus
capturing our uncertainty about their actual values. One important
difference between such a result and the result of a classical
statistical analysis is that classical confidence intervals (CIs) have a
rather complicated and somewhat confusing interpretation, which has
little to do with the posterior probability distribution.\footnote{Here
  are a few usual problems. CIs are often mistakenly interpreted as
  providing the probability that a resulting confidence interval
  contains the true value of a parameter. CIs bring confusion also with
  regard to precision, it is a common mistake to interpret narrow
  intervals as the ones corresponding to a more precise knowledge.
  Another fallacy is to associate CIs with likelihood and stating that
  values within a given interval are more probable than the ones outside
  it. The theory of confidence intervals does not support the above
  interpretations. CIs should be plainly interpreted as a result of
  certain procedure (there are many ways to obtain CIs from a given set
  of data) that will in the long run contain the true value if the
  procedure is performed a fixed amount of times. For a nice survey and
  explanation of these misinterpretations, see {[}17{]}. For a
  psychological study of the occurrence of such misinterpretations, see
  {[}9{]}. In this study, 120 researchers and 442 students were asked to
  assess the truth value of six false statements involving different
  interpretations of a CI. Both researchers and students endorsed, on
  average, more than three of these statements.}

@@

There are a few related papers, whose discussion goes beyond the scope
of this paper:

\begin{itemize}
\item
  {[}21{]} employ Bayesian methods to estimate uncertainty in NLP tasks,
  but they apply their Bayesian Neural Networks-based method to
  sentiment analysis or named entity recognition, not to bias.
\item
  {[}3{]} correctly argues that a bias estimate should not be expressed
  as a single number without taking into account that the estimate is
  made using a sample of data and therefore has intrinsic uncertainty.
  The authors suggest using Bernstein bounds to gauge the uncertainty in
  terms of confidence intervals. We do not discuss this approach
  extensively, as we think that confidence intervals are quite
  problematic for several reasons, among others the confusing
  interpretability. We do not think that Bernstein bounds provide the
  best solution to the problem. Applying this method to a popular
  WinoBias dataset leads to the conclusion that more than 11903 samples
  are needed to claim a 95\% confidence interval for a bias estimate.
  This amount vastly exceeds the existing word lists for bias
  estimation. We propose a more realistic Bayesian method. Our
  conclusion is still that the word lists are sometimes too small, but
  at least they allow for gauging uncertainty as we go on to improve our
  methodology and extend the lists gradually.
\item
  {[}20{]} criticize some existing bias metrics such as \textsf{MAC} or
  \textsf{WEAT} on the grounds of them not satisfying some general
  formal principles, such as magnitude-comparability, and they propose a
  modification, called \textsf{SAME}.
\item
  {[}14{]} develop a generalization of \textsf{WEAT} meant to apply to
  sets of sentences, \textsf{SEAT}, which basically applies
  \textsf{WEAT} to vector representations of sentences. The authors,
  however, still pre-average and play the game of finding a
  single-number metric, so our remarks apply.
\item
  {[}8{]} introduce the Contextualized Embedding Association Test
  Intersectional, meant to apply to dynamic word embeddings and,
  importantly develop methods for intersectional bias detection. The
  measure is a generalization of the WEAT method. The authors do inspect
  a distribution of effect sizes that arises from the consideration of
  various possible contexts, but they continue to standardize the
  difference in averaged means and use a single-number summary: the
  weighted mean of the effect sizes thus understood. The method,
  admittedly, deserves further evaluation which goes beyond the scope of
  this paper.
\end{itemize}

To summarize, a Bayesian data analysis with hierarchical models of
cosine distances between protected words, control group words, and
stereotypical attributes provides more modest and realistic assessment
of the uncertainty involved. It reveals that much complexity is hidden
when one instead chases single bias metrics present in the literature.

After introducing the method, we applied it to multiple word embeddings
and results of supposed debiasing, putting forward some general
observations that are not exactly in line with the usual picture painted
in terms of \textsf{WEAT} or \textsf{MAC} (and the problem generalizes
to any approach that focuses on chasing a single numeric metric): the
word list sizes and sample sizes used in the studies are usually small.
Posterior density intervals are fairly wide. Often the differences
between associated, different and neutral human predicates, are not very
impressive. Also, a preliminary inspection suggests that the
desirability of changes obtained by the usual debiasing methods is
debatable. The tools that we propose, however, allow for a more
fine-grained and multi-level evaluation of bias and debiasing in
language models without losing modesty about the uncertainties involved.

The short general and somewhat disappointing lesson here is this: things
are complicated. Instead of chasing single-number metrics, we should
rather devote attention to more nuanced analysis.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Caliskan2017semanticsBiases}{}}%
\CSLLeftMargin{{[}2{]} }%
\CSLRightInline{Aylin Caliskan, Joanna J. Bryson, and Arvind Narayanan.
2016. Semantics derived automatically from language corpora necessarily
contain human biases. \emph{CoRR} abs/1608.07187, (2016). Retrieved from
\url{http://arxiv.org/abs/1608.07187}}

\leavevmode\vadjust pre{\hypertarget{ref-Ethayarajh2020Bernstein}{}}%
\CSLLeftMargin{{[}3{]} }%
\CSLRightInline{Kawin Ethayarajh. 2020. Is your classifier actually
biased? Measuring fairness under uncertainty with bernstein bounds.
\emph{CoRR} abs/2004.12332, (2020). Retrieved from
\url{https://arxiv.org/abs/2004.12332}}

\leavevmode\vadjust pre{\hypertarget{ref-Garg2017hundredYears}{}}%
\CSLLeftMargin{{[}4{]} }%
\CSLRightInline{Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James
Zou. 2017. Word embeddings quantify 100 years of gender and ethnic
stereotypes. \emph{Proceedings of the National Academy of Sciences} 115,
(November 2017).
DOI:https://doi.org/\href{https://doi.org/10.1073/pnas.1720347115}{10.1073/pnas.1720347115}}

\leavevmode\vadjust pre{\hypertarget{ref-Garg2018years}{}}%
\CSLLeftMargin{{[}5{]} }%
\CSLRightInline{Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James
Zou. 2018. Word embeddings quantify 100 years of gender and ethnic
stereotypes. \emph{Proceedings of the National Academy of Sciences} 115,
16 (April 2018), E3635--E3644.
DOI:https://doi.org/\href{https://doi.org/10.1073/pnas.1720347115}{10.1073/pnas.1720347115}}

\leavevmode\vadjust pre{\hypertarget{ref-Gonen2019lipstick}{}}%
\CSLLeftMargin{{[}6{]} }%
\CSLRightInline{Hila Gonen and Yoav Goldberg. 2019. Lipstick on a pig:
{D}ebiasing methods cover up systematic gender biases in word embeddings
but do not remove them. In \emph{Proceedings of the 2019 conference of
the north {A}merican chapter of the association for computational
linguistics: Human language technologies, volume 1 (long and short
papers)}, Association for Computational Linguistics, Minneapolis,
Minnesota, 609--614.
DOI:https://doi.org/\href{https://doi.org/10.18653/v1/N19-1061}{10.18653/v1/N19-1061}}

\leavevmode\vadjust pre{\hypertarget{ref-gordon2012reporting}{}}%
\CSLLeftMargin{{[}7{]} }%
\CSLRightInline{Jonathan Gordon and Benjamin Durme. 2013. Reporting bias
and knowledge acquisition. In \emph{AKBC 2013 - Proceedings of the 2013
Workshop on Automated Knowledge Base Construction, Co-located with CIKM
2013}, 25--30.
DOI:https://doi.org/\href{https://doi.org/10.1145/2509558.2509563}{10.1145/2509558.2509563}}

\leavevmode\vadjust pre{\hypertarget{ref-Guo2021CEAT}{}}%
\CSLLeftMargin{{[}8{]} }%
\CSLRightInline{Wei Guo and Aylin Caliskan. 2021. Detecting emergent
intersectional biases: Contextualized word embeddings contain a
distribution of human-like biases. In \emph{Proceedings of the 2021
{AAAI}/{ACM} conference on {AI}, ethics, and society}, {ACM}.
DOI:https://doi.org/\href{https://doi.org/10.1145/3461702.3462536}{10.1145/3461702.3462536}}

\leavevmode\vadjust pre{\hypertarget{ref-Hoekstra2014Misinterpretation}{}}%
\CSLLeftMargin{{[}9{]} }%
\CSLRightInline{Rink Hoekstra, Richard D. Morey, Jeffrey N. Rouder, and
Eric-Jan Wagenmakers. 2014. Robust misinterpretation of confidence
intervals. \emph{Psychonomic Bulletin \& Review} 21, 5 (October 2014),
1157--1164.
DOI:https://doi.org/\href{https://doi.org/10.3758/s13423-013-0572-3}{10.3758/s13423-013-0572-3}}

\leavevmode\vadjust pre{\hypertarget{ref-JohnsonValueFree}{}}%
\CSLLeftMargin{{[}10{]} }%
\CSLRightInline{Gabbrielle Johnson. forthcoming. Are algorithms
value-free? Feminist theoretical virtues in machine learning.
\emph{Journal Moral Philosophy} (forthcoming).}

\leavevmode\vadjust pre{\hypertarget{ref-kruschke2015bayesian}{}}%
\CSLLeftMargin{{[}11{]} }%
\CSLRightInline{John Kruschke. 2015. \emph{Doing {B}ayesian data
analysis (second edition)}. Academic Press, Boston.}
